---
title: "An opinionated comparison of model-based and classical ordination methods"
institute: "Chair of Plant Ecology, University of Bayreuth"
author: "Bert van der Veen"
output: 
  beamer_presentation:
    toc: false
    slide_level: 2
    latex_engine: lualatex
    includes:
      in_header: header.tex
urlcolor: orange
editor_options: 
  chunk_output_type: console
header-includes:
  - \usepackage{ulem}
  - \usepackage{pgf}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \tcbuselibrary{skins}
  - \usepackage{tikz}
  - \usepackage{xcolor}
  - \usepackage{graphicx}
  - \usetikzlibrary{calc, fit, positioning,arrows,shapes,tikzmark,spy,shadows.blur,matrix, arrows.meta}
  - \tikzset{>=stealth}
  - \tikzstyle{na} = [baseline=-.5ex]
  - \usepackage{multicol}
  - \usepackage{hyperref}
  - \newcommand{\highlight}[2]{\colorbox{#1!17}{$\vphantom{(^\top}#2$}}
---

```{r setup, include=FALSE}
library(knitr)

default_source_hook <- knit_hooks$get('source')
default_output_hook <- knit_hooks$get('output')

knit_hooks$set(
  source = function(x, options) {
    paste0(
      "\n::: {.codebox data-latex=\"\"}\n\n",
      default_source_hook(x, options),
      "\n\n:::\n\n")
  }
)

knit_hooks$set(
  output = function(x, options) {
    paste0(
      "\n::: {.codebox data-latex=\"\"}\n\n",
      default_output_hook(x, options),
      "\n\n:::\n\n")
  }
)

knitr::opts_chunk$set(echo = TRUE)
```

# Outline

- Part I: Overview
  - Model-based ordination
  - Compared to classical methods
- Part II: Examples
  - Cover classes
  - Biomass
  - Continuous cover with complete presence/absence
- Part III: Summary
  - Contrast to classical methods

Slides and code: \url{https://github.com/BertvanderVeen/NINA}


# Part I: Overview

## Limitations of classical multivariate analysis

Classical ordination methods are limited:

- Loose framework of separate methods
- No explicit model structure
- Cannot accommodate data properties
- Strong assumptions about data properties and structure
- Assumptions are not verifiable
- No uncertainty
- Lacks predictive capacity

## Model-based ordination is the next step

\begin{columns}
\begin{column}{0.5\textwidth}
Classical ordination stops at the biplot.
\textcolor{red}{NMDS and distance-based methods traded process for pattern.}

\vspace*{\baselineskip}

With model-based ordination, the biplot is only the beginning.

\vspace*{\baselineskip}

It is also where ordination was headed \tiny (ter Braak and Prentice 1988, ter Braak 1985, Ihm and van Groenewoud 1975, Gauch et al. 1974) \normalsize prior to Minchin (1987) popularizing NMDS

\end{column}
\begin{column}{0.5\textwidth}
```{r unc_ordi, echo=FALSE, message=FALSE, warning=FALSE, fig.height=5, fig.width=4, cache=TRUE, fig.align='center'}
library(gllvm)
invisible(TMB::openmp(11, autopar = TRUE))
data(spider, package = "mvabund")
fit_unc <- gllvm(spider$abund, num.lv = 2, family = "negative.binomial", seed = 1, quadratic = TRUE)
ordiplot(fit_unc, symbols = TRUE, s.colors = "grey50",
         main = "Unimodal unconstrained ordination", cex.main = 0.9, biplot = TRUE, xlim = c(-4,4), ylim = c(-3,3))
```
\end{column}
\end{columns}


## What is model-based ordination

\vspace{0.5cm}
\begin{equation*}
    \tikzmarknode[fill=teal!20, inner sep=2pt, rounded corners=2pt]{dist}{y_{ij}} \sim
    \tikzmarknode[fill=teal!20, inner sep=2pt, rounded corners=2pt]{distr}{f(\mu_{ij}, \phi_j)}
\end{equation*}
\only<2->{
\begin{equation*}
    \tikzmarknode[fill=blue!20, inner sep=2pt, rounded corners=2pt]{link}{g(\mu_{ij})} =
    \only<3->{\tikzmarknode[fill=red!20, inner sep=2pt, rounded corners=2pt]{alpha}{\alpha_i} +}
    \tikzmarknode[fill=orange!20, inner sep=2pt, rounded corners=2pt]{beta0}{\beta_{0j}}
    \only<4->{+ \tikzmarknode[fill=green!20, inner sep=2pt, rounded corners=2pt]{env}{\mathbf{x}_i^\top \symbf{\beta}_j} +
    \tikzmarknode[fill=cyan!20, inner sep=2pt, rounded corners=2pt]{specre}{\mathbf{z}_i^\top \mathbf{b}_j}}
    + \tikzmarknode[fill=purple!20, inner sep=2pt, rounded corners=2pt]{latent}{\mathbf{u}_i^\top \symbf{\gamma}_j}
\end{equation*}
}
\only<3->{
\begin{equation*}
    \alpha_i = \tikzmarknode[fill=red!10, inner sep=2pt, rounded corners=2pt]{alphafixed}{\mathbf{w}_i^\top \symbf{\delta}} +
    \tikzmarknode[fill=red!30, inner sep=2pt, rounded corners=2pt]{alphare}{a_i}, \quad
    \mathbf{a} \sim \mathcal{N}(\mathbf{0}, \symbf{\Sigma}^a), \quad
    \mathrm{vec}(\mathbf{B}) \sim \mathcal{N}\!\left(\mathbf{0},\, \symbf{\Sigma}^r \otimes \symbf{\Sigma}\right)
\end{equation*}
}

\begin{tikzpicture}[
    overlay, remember picture,
    >={Stealth[scale=1]},
    annot/.style={font=\footnotesize\sffamily, align=center}
]
    \only<1>{
        \node[annot, text=teal!80!black, above=0.8cm of distr] (t_distr) {Distribution};
        \draw[->, teal!80!black] (t_distr) -- (distr.north);
    }
    \only<2->{
        \node[annot, text=blue!80!black, above left=0.7cm and 0.2cm of link] (t_link) {Link\\function};
        \draw[->, blue!80!black] (t_link.south) to[out=-80, in=150] (link.west);

        \node[annot, text=orange!80!black, below=2.0cm of beta0] (t_beta0) {Species\\intercepts};
        \draw[->, orange!80!black] (t_beta0) -- (beta0.south);

        \node[annot, text=purple!80!black, below right=0.8cm and 0.8cm of latent] (t_latent) {Ordination};
        \draw[->, purple!80!black] (t_latent.west) to[out=170, in=-50] (latent.south);
    }
    \only<3->{
        \node[annot, text=red!80!black, above left=1.8cm and 1.5cm of alpha] (t_alpha) {Site\\intercepts};
        \draw[->, red!80!black] (t_alpha.east) to[out=0, in=130] (alpha.north);

        \node[annot, text=red!60!black, below=0.5cm of alphafixed] (t_alphafixed) {Fixed effects};
        \draw[->, red!60!black] (t_alphafixed) -- (alphafixed.south);

        \node[annot, text=red!80!black, below=1.0cm of alphare] (t_alphare) {Random effect};
        \draw[->, red!80!black] (t_alphare) -- (alphare.south);
    }
    \only<4->{
        \node[annot, text=green!60!black, above right=1.8cm and 0.5cm of env] (t_env) {Environmental predictors\\(with species coefficients)};
        \draw[->, green!60!black] (t_env.south) to[out=-100, in=60] (env.north);

        \node[annot, text=cyan!60!black, below=1.2cm of specre] (t_specre) {Species\\random effects};
        \draw[->, cyan!60!black] (t_specre) -- (specre.south);
    }
\end{tikzpicture}

## Key advantages of Model-based ordination

\footnotesize

- \textbf{Distribution-based}: no distances or data transformations; direct connection to data is maintained
- \textbf{Random effects}: integrates ordination with mixed-effects modeling for hierarchical designs
- \textbf{Diagnostics and inference}: enables assumption checking, model evaluation, uncertainty quantification
- \textbf{Predictive capacity}: enables (multivariate) forecasting
- \textbf{Diversity estimation}: facilitates inference on diversity in the same framework

# Part II: Examples

Each example steps further from what classical ordination can do:

\begin{enumerate}
\item Cover classes from Skabbholmen, Sweden
\begin{itemize}
\item design-aware ordination: random effects change interpretation
\end{itemize}
\item Biomass data from the Waddensea, Germany
\begin{itemize}
\item ordination and prediction are the same model
\end{itemize}
\item Cover data from Rothamsted, United Kingdom
\begin{itemize}
\item predicting an emergent community property: species richness
\end{itemize}
\end{enumerate}

## Example 1: Skabbholmen, Sweden

- **Sites**: 64 unique plots across 4 transects
- **Species**: 49 vascular plant species with cover data on a 5-degree Hult-Sernander-Du Rietz scale
- **Predictors**: 
  - Year (1978 vs 1984) - representing temporal change
  - Elevation (centimeters) - representing long-term land uplift
- **Data type**: Cover classes
- **Analysis**: Cumulative probit/logit model with 2 informed LVs

Previously analyzed by [ter Braak (1987)](https://link.springer.com/article/10.1007/BF00038688)
\newline
Originally by [Cramer and Hytteborn (1987)](https://link.springer.com/article/10.1007/BF00038697)

## Example 1: Does design matter for ordination?

Three versions of the same model — differing only in how transect structure is handled:

- **No transect RE**: latent variables must absorb any spatial clustering between transects
- **Species-common transect RE**: one intercept shift per transect, shared across all species
- **Species-specific transect RE**: each species gets its own transect random effect

```{r skabb_mod, warning=F, error=F, message=F, echo=F, message=F, warning=F, cache = TRUE}
# with and without row effect: confounding

library(gllvm)
invisible(TMB::openmp(11,autopar=TRUE))

data("Skabbholmen")
Y <- Skabbholmen$Y[,order(colSums(ifelse(Skabbholmen$Y==0,0,1)),decreasing=TRUE)]
X <- Skabbholmen$X
# Fit the model
model <- gllvm(y = Y, X = X, studyDesign = X,
    lv.formula = ~scale(Elevation) + I((Year-1978)/6),
    num.lv.c = 2,
    family = "ordinal",
    zeta.struc="common",
    row.eff=~(1|transectID),
    quadratic = "LV",
    n.init = 1,link="logit", seed = 9705, jitter.var=0.1
    )
model2 <- gllvm(y = Y, X = X, 
    lv.formula = ~scale(Elevation) + I((Year-1978)/6),
    num.lv.c = 2,
    family = "ordinal",
    zeta.struc="common",
    quadratic = "LV",
    n.init = 1,link="logit", seed = 9705,jitter.var=0.1,
    )
model3 <- gllvm(y = Y, X = X,
    lv.formula = ~scale(Elevation) + I((Year-1978)/6),
    num.lv.c = 2,
    family = "ordinal",
    zeta.struc="common",
    formula=~(1|transectID),
    quadratic = "LV",
    link="logit", seed = 5922 # because this takes quite long; ran it 100 times to get the seed
    )
```

## Example 1: Random effects change the ordination

```{r ordiplot_skabb_opt, warning=F, error=F, message=F, echo=F, fig.width=9, fig.height=3.2, out.width="\\linewidth"}
#knitr::load_cache("skabb_mod",path="/home/bertv/github/NINA/ModelvsClassic_cache/beamer/")
library(ggplot2)

# Rotate each model so its elevation loading has zero first-dimension component
get_data <- function(object) {
  LV   <- gllvm::getLV(object)
  coef <- object$params$LvXcoef
  sds  <- apply(object$lv.X.design, 2, sd)
  sds[sds == 0] <- 1
  coef <- coef / sds
  # rotation angle: set elevation's LV1 component to zero (point straight up)
  elev_row <- grep("elev", rownames(coef), ignore.case = TRUE)[1]
  ev    <- coef[elev_row, ]
  theta <- pi/2 - atan2(ev[2], ev[1])
  R     <- matrix(c(cos(theta), sin(theta), -sin(theta), cos(theta)), 2, 2, byrow = TRUE)
  list(LV = LV %*% R, coef = coef %*% R)
}

d2 <- get_data(model2)
d1 <- get_data(model)
d3 <- get_data(model3)

# Tidy site-scores data frame
model_labels <- c("No transect RE", "Species-common transect RE", "Species-specific transect RE")
sites <- do.call(rbind, Map(function(d, lab) {
  data.frame(LV1 = d$LV[,1], LV2 = d$LV[,2], transect = X$transectID, model = lab)
}, list(d2, d1, d3), model_labels))
sites$model <- factor(sites$model, levels = model_labels)

# Scale arrows to 80% of the global LV range so all panels share the same scale
global_sc <- 0.8 * max(abs(c(d1$LV, d2$LV, d3$LV)))

make_arrows <- function(d, lab) {
  coef_sc <- d$coef / max(abs(d$coef)) * global_sc
  idx     <- apply(coef_sc, 1, function(x) max(abs(x)) > 0.001)
  coef_sc <- coef_sc[idx, , drop = FALSE]
  data.frame(x1 = coef_sc[,1], y1 = coef_sc[,2],
             label = rownames(coef_sc), model = lab, stringsAsFactors = FALSE)
}
arrows <- do.call(rbind, Map(make_arrows, list(d2, d1, d3), model_labels))
arrows$model <- factor(arrows$model, levels = model_labels)

n_tr <- nlevels(X$transectID)
ggplot(sites, aes(LV1, LV2, colour = transect, shape = transect)) +
  geom_point(size = 2) +
  geom_segment(data = arrows,
               aes(x = 0, y = 0, xend = x1, yend = y1),
               arrow = arrow(length = unit(0.15, "cm"), type = "closed"),
               colour = "red", linewidth = 0.8, inherit.aes = FALSE) +
  geom_text(data = arrows,
            aes(x = x1 * 1.15, y = y1 * 1.15, label = label),
            colour = "red", size = 4, inherit.aes = FALSE) +
  scale_colour_manual(values = palette()[seq_len(n_tr)]) +
  scale_shape_manual(values = seq_len(n_tr) + 14L) +
  facet_wrap(~model, nrow = 1) +
  labs(x = "Latent variable 1", y = "Latent variable 2",
       colour = "Transect", shape = "Transect") +
  theme_classic(base_size = 14) +
  theme(legend.position = "bottom",
        strip.background = element_blank(),
        panel.spacing = unit(0.8, "lines"))
```

- Without a transect RE, transect effect are confounded with the environmental space: the axes reflect clustering
- Adding the RE removes the confounding, the latent variables align with the elevation and year effects
- The choice of RE structure is a modelling decision — one that classical ordination cannot make

## Example 2: Wadden Sea, Germany

\begin{figure}[h]
\centering
\includegraphics[width=\linewidth,height=0.7\textheight,keepaspectratio]{Dewenterfig1.jpg}
\caption{nioz.nl}
\end{figure}

## Example 2: Wadden Sea, Germany

- **Sites**: 3 sites at 3 transects on each of 3 islands
- **Species**: 58 macrozoobenthos species
- **Predictors**: 
  - Season - representing temporal variation
  - TOC - total organic carbon in sediment
  - Elevation - elevation of sediment surface
  - Mud - percentage of mud in sediment
  - Chlorophyll a - primary production
  - DIN -  dissolved inorganic nitrogen
  - RDP - respiration rate of sediment
- **Data type**: Biomass data
- **Analysis**: Tweedie log-linear model with 2 constrained latent variables

Originally by [Dewenter et al. (2023)](https://onlinelibrary.wiley.com/doi/full/10.1002/ece3.10815)

```{r wadden_example, echo=FALSE, message=FALSE, warning=FALSE, cache = TRUE}
library(gllvm)
invisible(TMB::openmp(11,autopar=TRUE))
# Load the Wadden Sea data (macrozoobenthos abundance data)
Y <- read.csv("waddenY2.csv")[,-c(1:2)]
Y <- Y[, colSums(ifelse(Y==0,0,1))>2]  # Remove species with less than 3 non-zero values
X <- read.csv("waddenX.csv")
X <- X[,!apply(X,2,anyNA)]  # Remove columns with NA values
X[,unlist(lapply(X,is.numeric))] <- scale(X[,unlist(lapply(X,is.numeric))])  # Scale numeric variables

# Unconstrained ordination (simple model)
#uord1 <- gllvm(Y, num.lv = 2, family = "tweedie", Power = NULL)

# Constrained ordination with environmental variables
# cord1 <- gllvm(Y, X = X, 
#               num.RR = 2, lv.formula = ~season+TOC+I(TOC^2)+elevation+I(elevation^2)+silt_clay+I(silt_clay^2)+Chl.a+I(Chl.a^2)+DIN+I(DIN^2)+RDP+I(RDP^2), randomB = "LV", 
#               family = "tweedie", Power = NULL, disp.formula = rep(1,ncol(Y)), seed = 1172) # after 200 iterations
# added fixed row effects for lfexibilityf
cord1 <- gllvm(Y, X = X, 
               num.RR = 2, lv.formula = ~season+TOC+I(TOC^2)+elevation+I(elevation^2)+silt_clay+I(silt_clay^2)+Chl.a+I(Chl.a^2)+DIN+I(DIN^2)+RDP+I(RDP^2), randomB = "LV", 
               family = "tweedie", Power = NULL, disp.formula = rep(1,ncol(Y)), row.eff = ~season+TOC+I(TOC^2)+elevation+I(elevation^2)+silt_clay+I(silt_clay^2)+Chl.a+I(Chl.a^2)+DIN+I(DIN^2)+RDP+I(RDP^2), studyDesign = X, seed = 1) 
# Plot the constrained ordination
# gllvm::ordiplot(cord1, symbols=TRUE, pch=X$island, 
#                 s.colors = as.character(factor(X$island,labels  = c("blue", "green", "black"))), 
#                 arrow.scale=0, s.cex = 2, cex.main = 2, cex.axis = 2, cex.lab = 2, 
#                main = "Constrained ordination of Wadden Sea data")

```

## Example 2: Constrained ordination

```{r wadden_ordiplot, echo = FALSE, warning=FALSE, message=FALSE, fig.height = 4, fig.width = 4}
library(ggplot2)

# Extract site scores (latent variables) and predictor coefficients
LV <- gllvm::getLV(cord1)
coefs <- cord1$params$LvXcoef
sds <- apply(cord1$lv.X.design, 2, sd)
sds[sds == 0] <- 1
coefs <- coefs / sds

# Keep only main effects (no quadratic terms) for cleaner arrows
main_idx <- !grepl("^I\\(", rownames(coefs))
coefs_main <- coefs[main_idx, , drop = FALSE]

# Scale arrows relative to site score range
sc <- 0.8 * max(abs(LV))
coefs_sc <- coefs_main / max(abs(coefs_main)) * sc

arrow_df <- data.frame(
  x1 = coefs_sc[, 1], y1 = coefs_sc[, 2],
  label = rownames(coefs_sc)
)

site_df <- data.frame(LV1 = LV[, 1], LV2 = LV[, 2], season = X$season)

ggplot(site_df, aes(LV1, LV2, colour = season)) +
  geom_point(size = 1.5) +
  geom_segment(data = arrow_df,
               aes(x = 0, y = 0, xend = x1, yend = y1),
               arrow = arrow(length = unit(0.15, "cm"), type = "closed"),
               colour = "red", linewidth = 0.4, inherit.aes = FALSE) +
  geom_text(data = arrow_df,
            aes(x = x1 * 1.15, y = y1 * 1.15, label = label),
            colour = "red", size = 3, inherit.aes = FALSE) +
  labs(x = "Latent variable 1", y = "Latent variable 2", colour = "Season") +
  theme_classic(base_size = 12) +
  theme(legend.position = "bottom")
```

## Example 2: Variance partitioning

```{r wadden_example_VPplot, echo = FALSE, warning=FALSE, message=FALSE, fig.height = 5, out.height = "70%",out.width = "100%",fig.width = 9,fig.align= "center"}
plot(gllvm::VP(cord1), las =2, args.legend = list(cex = 0.7, bg = "white"), col = hcl.colors(5, "Dark 3"), lwd = 0.5,xlab=NA,main=NA,mar=c(0,0,0,0))
```

Variance partitioning attributes a large share of explained variance to TOC — suggesting it is an important driver.

## Example 2: Species-specific results

```{r wadden_example_SPplot, echo = FALSE, warning=FALSE, message=FALSE,out.width = "49%", fig.width=4, fig.height=4.5}
gllvm::randomCoefplot(cord1,which.Xcoef="TOC", mar = c(4,5,0,1))
gllvm::randomCoefplot(cord1,which.Xcoef="I(TOC^2)",y.label = FALSE,  mar = c(4,0,0,1))
```

Yet species-specific TOC effects are highly uncertain and vary widely across species — variance alone is not enough.

## Example 2: Species-level predictions

```{r preds, echo = FALSE, cache = TRUE}
newX = data.frame(elevation = rep(seq(min(X$elevation), max(X$elevation), length.out = 100),times=2), silt_clay = mean(X$silt_clay),
                                              season = factor(rep(c("spring","summer"), each = 100)),
                                              TOC = mean(X$TOC),
                                              Chl.a = mean(X$Chl.a), DIN = mean(X$DIN), RDP = mean(X$RDP))
                                              
preds = predict(cord1, newX = newX, se.fit = 20e3)
                                              
```

```{r wadden_example_plot, echo = FALSE, warning=FALSE, message=FALSE, fig.height = 9, fig.width = 14, fig.align = "center"}
spp_names <- colnames(cord1$y)   # change to readable labels here if needed

plotdat_long <- data.frame(
  elevation = rep(newX$elevation, times = length(spp_names)),
  season    = rep(newX$season,    times = length(spp_names)),
  species   = factor(rep(spp_names, each = nrow(newX)), levels = spp_names),
  fit       = as.vector(preds$fit),
  lwr       = as.vector(preds$lower),
  upr       = as.vector(preds$upper)
)

library(ggplot2)
ggplot(plotdat_long, aes(elevation, fit, colour = season, fill = season)) +
  geom_ribbon(aes(ymin = lwr, ymax = upr, fill = season), alpha = 0.3, colour = NA) +
  geom_line(linewidth = 0.2) +
  facet_wrap(~ species, scales = "free_y") +
  labs(y = "log(Predicted response)", x = "Elevation",
       colour = "Season", fill = "Season") +
  theme_classic(base_size = 12) +
  theme(legend.position = "bottom", strip.background = element_blank(),, axis.title.x = element_text(size = 23), axis.title.y  = element_text(size = 23),
        panel.spacing = unit(0.8, "lines")) +
  geom_rug(aes(x = elevation), data = X, inherit.aes = FALSE, length = unit(0.1, "npc"))
```

## Example 3: Rothamsted, United Kingdom

- **Sites**: 16 plots, each with subplots
- **Species**: 57 vascular plant species
- **Predictors**: 
  - Fertilizer treatment
    - Phosphorus, Potassium, Nitrogen, Liming, and so on
- **Data type**: Continuous cover with absence and presence
- **Analysis**: Ordered beta model with two constrained latent variables

Originally studied by a.o. [Crawley et al. (2005)](https://oro.open.ac.uk/2295/1/Crawley_etal_2005.pdf)

## Example 3: From ordination to diversity

Species richness is not modelled directly — it emerges from the joint species model:

- No separate richness model is needed — richness emerges from the joint species model for any data type
- Uncertainty propagates naturally from species-level predictions, avoiding distributional assumptions on richness directly

\pause

Both the species responses and the community-level summary are coherently estimated in one step.

## Example 3: Species richness predictions

```{r parkgrass_analysis, echo=FALSE, message=FALSE, warning=FALSE, cache = TRUE}
library(readxl)

Y <- read.csv("/home/bertv/Dropbox/SR_regression/park-grass/parkgrass.csv")
soil <- read_xlsx("/home/bertv/Dropbox/SR_regression/park-grass/01-PGsoilpH/Park Grass soil pH 1876-2023 - data-PG-soil-pH-1876-2023-01.xlsx", sheet = 3)
dat <-  merge(Y,soil,by.x=c("plot_id","sample_year"),by.y =  c("plot_id","harvest_year"))
dat <- dat[!is.na(dat$pH_soil),]
Y <- dat[,28:95]
X <- dat[,-c(28:95)]
Y <- Y[,colSums(Y,na.rm=TRUE)!=0]
# Two NA rows for some reason
X <- X[-which(rowSums(Y,na.rm=TRUE)==0),]
Y <- Y[-which(rowSums(Y,na.rm=TRUE)==0),]
Y <- Y[,colSums(ifelse(Y==0,0,1))>3]
library(gllvm)
invisible(TMB::openmp(11,autopar=TRUE))
X$pH_soil.s <- scale(X$pH_soil)
setMap = numeric(ncol(Y*2))
setMap[seq_len(ncol(Y))*2] <- NA
setMap[seq_len(ncol(Y))*2-1] <- 1:ncol(Y)

X$treatment <- relevel(as.factor(X$treatment), ref = "Nil")
# contrasts(X$treatment) <- contr.sum(nlevels(X$treatment))

setMap = numeric(ncol(Y*2))
setMap[seq_len(ncol(Y))*2] <- NA
setMap[seq_len(ncol(Y))*2][which(apply(Y,2,function(x)sum(x==100))>0)]<-ncol(Y)+1
setMap[seq_len(ncol(Y))*2-1] <- 1:ncol(Y)

model <- gllvm(y = Y/100, X = X[, c("pH_soil.s", "treatment"), drop = FALSE], family = "orderedBeta", num.RR = 2, lv.formula = ~(pH_soil.s + I(pH_soil.s^2) + I(pH_soil.s^3)) * treatment, studyDesign = X,
               row.eff = ~(pH_soil.s + I(pH_soil.s^2) + I(pH_soil.s^3)) * treatment,
               randomB = "LV", zeta.struc = "common", link = "logit", disp.formula = rep(1, ncol(Y)), zetacutoff = c(-2, 5), seed = 1222) # after 84 iterations
```

```{r park_grasss_SR_calc, echo = FALSE,cache=TRUE}
# Make predictions for all treatments
x <- aggregate(X$pH_soil.s,by=list(X$treatment), FUN = range)
newX <- do.call(rbind,sapply(1:19,function(i)data.frame(pH_soil.s = seq(x[[2]][i,1],x[[2]][[i,2]],length.out=50),treatment = x[[1]][i]), simplify=FALSE))
newX$treatment <- relevel(newX$treatment,ref="Nil")
colnames(newX) = c("pH_soil.s","treatment")

invisible(gllvm:::set_omp_threads(11))
test<- gllvm::predictSR(model, newX = newX)
``` 

```{r parkgrass_plot, echo=FALSE, message=FALSE, warning=FALSE, out.height = "85%", fig.align = "center"}

# Prepare data frames for plotting
df_pred <- data.frame(
  pH = newX$pH_soil.s * sd(X$pH_soil) + mean(X$pH_soil),
  y = test$expected$fit,
  lower = test$expected$lower,
  upper = test$expected$upper,
  treatment = factor(newX$treatment, levels = unique(X$treatment))
)

df_obs <- data.frame(
  pH = X$pH_soil,
  y = rowSums(ifelse(model$y == 0, 0, 1)),
  treatment = factor(X$treatment, levels = unique(X$treatment))
)

# Define a logical order for your treatments
treatment_order <- c(
  "Nil", "P", "K", "FYM/FM", "(FYM/FM)", "N1", "N*1", "N2+P", 
  "PNaMg", "PKNaMg", "N1+PKNaMg", "N*1+PKNaMg", "N2+PNaMg", 
  "N2+PKNaMg", "N*2+PKNaMg", "(N*2)PKNaMg", "(N2)PKNaMg", 
  "N3+PKNaMg", "N3+PKNaMgSi" 
)

df_pred$treatment <- factor(newX$treatment, levels = treatment_order)
df_obs$treatment <- factor(X$treatment, levels = treatment_order)

library(ggplot2)

# Plot with facets
ggplot() +
  geom_line(data = df_pred, aes(x = pH, y = y), linewidth = 0.8, color = "black") +
  geom_ribbon(data = df_pred, aes(ymin = lower, ymax = upper, x = pH), fill = "gray", alpha = 0.5) +
  geom_point(data = df_obs, aes(x = pH, y = y), shape = 16, size = 1, color = "red") +
  facet_wrap(~treatment) +
  labs(x = "Soil pH", y = "Species richness") +
  theme_classic(base_size=16)
```

# Part III: Summary

## What the examples show \footnotesize \newline and what classical ordination cannot do

\begin{columns}
\begin{column}{0.48\textwidth}
\begin{enumerate}
\item \textbf{Skabbholmen}: Ignoring hierarchical design confounds the ordination
\item \textbf{Wadden Sea}: With model-based ordination species-level forecasting comes for free
\item \textbf{Park Grass}: When the species data is modelled, community-level metrics follow
\end{enumerate}
\end{column}
\begin{column}{0.48\textwidth}
\only<2->{
\textbf{Classical ordination:}
\begin{enumerate}
\item No hierarchical design: RDA/CCA cannot incorporate random effects
\item Stop at the biplot: there is no prediction or uncertainty
\item Severs the data connection: you cannot extract diversity
\end{enumerate}
}
\end{column}
\end{columns}

\only<3->{
Pattern visualization → design-aware inference → multivariate prediction → community-level forecasting.
}

## Summary

Model-based ordination is not more complex, but more complete:

- All data types handled in a single, coherent framework
- Non-linearity: what NMDS approximates via ranks, model-based ordination handles explicitly via the link function and model
- Rare species: data deficiency offset by partial pooling across species
- Assumptions are explicit, verifiable, and can be relaxed
- Uncertainty is propagated, not ignored
- Hierarchical designs are accommodated naturally
- Prediction is built in, not bolted on


## Future directions

- Diversity measures directly from multispecies models (van der Veen in prep., Many authors et al. in prep)
- Temporal; Rugstad et al. in prep.
- Spatial; Korhonen et al. in prep
- Better approximate likelihood methods
- Structured species' random effects
- Phylogenetic ordination, Hierarchical ordination  [(O'Hara and van der Veen preprint)](https://www.biorxiv.org/content/10.1101/2024.01.08.574701v1)

\pause

I haven't really spoken about species' correlation in a JSDM context, but it's here.